---
title: "AMMS Population Structure from Allele Frequencies Practical" 
author: "Izzy Routledge, Bob Verity"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    theme: readable
    highlight: tango
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, 
                      fig.align = 'center', fig.keep = 'all')
```

<!-- Here we style out button a little bit -->

```{=html}
<style>
.showopt {
background-color: #004c93;
color: #FFFFFF; 
width: 100px;
height: 20px;
text-align: center;
vertical-align: middle !important;
border-radius: 8px;
float:right;
}

.showopt:hover {
background-color: #dfe4f2;
color: #004c93;
}

</style>
```

# Dependencies for Practical {-}

Please copy and paste the below code chunk in it's entirety to your console to download R package libraries needed for this practical. If you are having trouble installing any of the R packages, please ask an instructor for a pre-loaded flash drive.

```{r, echo=T}
deps <- c("tidyverse", "genescaper", "MIPanalyzer","rmaverick")
deps <- !sapply(deps, function(x){x %in% installed.packages()[,1]})
if (any(deps)) {
  if (deps["tidyverse"]) {
    install.packages("tidyverse")
  }
  if (deps["genescaper"]) {
    devtools::install_github("mrc-ide/genescaper", ref = "v0.1.0")
  }
  if (deps["MIPanalyzer"]) {
    devtools::install_github("mrc-ide/MIPanalyzer", ref = "v1.0.0")
  }
  if (deps["rmaverick"]) {
    devtools::install_github("bobverity/rmaverick", ref = "v1.1.0")
  }
}

```

Now load all of those libraries into this session using the code chunk below. Please copy and paste it in its entirety.

```{r, echo=T}
library(tidyverse)
library(genescaper)
library(MIPanalyzer)
library(rmaverick)
```

Finally, source the additional functions that are needed for this practical by copy-pasting this function:

```{r, echo=T}
source("source_functions/pop_structure_utils.R")
```


# Intro to population structure and allele frequency metrics {-}


One of the most fundamental observations that we can make from genetic data is how the frequencies of different alleles vary over time and over space.

For loci where mutations confer drug resistance, allele frequencies tell us something with direct clinical implications - if an allele is at high frequencies then we might consider switching first-line drugs.

However, even for loci that have no direct impact on phenotype, and therefore are selectively neutral, allele frequencies can give us important information about what is going on in the parasite population. 

These loci will be influenced by factors such as prevalence, human and mosquito population size, and migration rates between different sub-populations. If we can correctly interpret these signals then we can use neutral allele frequencies to tell us things about population structure and connectivity that can be relevant for control purposes.

_**Definition:** An allele is the identity of a particular genetic locus or sequence that is inherited between parents and offspring._

_**Definition:** A locus is a fixed position on a chromosome where a particular genetic marker is located._

_**Definition:** A neutral allele is one which is not positively or negatively affecting an organism's fitness and undergoing selection._

### Overview of Data {-}

We will work with both simulated and real data in this practical. The simulated data will allow us to explore how allele frequencies change over time as a function of different underlying parameters - something that we can almost never do from messy real-world data. This will allow us to build up an intuition about what we would expect to see in real data. In the second half of the practical, we will use real-world datasetes including a large molecular inversion probe (MIP) dataset taken from DRC and surrounding countries, previously analysed by [Verity et al. 2020](https://pubmed.ncbi.nlm.nih.gov/32355199/). We will use this to explore population structure in various ways.

### Practical Goals {-}
By the end of this practical, you should be able to: 

- Describe how population size, mutation and migration rates affect the patterns that we expect to see in neutral allele frequency data
- Estimate metrics such as $F_{ST}$ and explain what they can tell us about population differentiation
- Use ordination techniques like PCA to explore patterns of population structure
- Use model-based techniques like *rmaverick* to detect population structure in a probability-driven way
- Appreciate how allele frequency methods can identify connectivity over large spatial scales and long temporal scales
- Explain how allele frequencies can give useful information, and also describe their limitations for malaria control purposes


-------------------------

# Population genetics through simulation {-}

First we will simulate a population and different frequencies of alleles. To do this we will use an R function called sim_freqs to generate allele frequencies.

## Focusing on a single subpopulation {-}

Let's consider a single subpopulation (e.g. village) where there is no barrier to gene flow, also known as a "deme". 

### Allele frequency changes in a single deme {-}

First we will simulate the allele frequencies of a population of 1000 individuals infected with malaria. Try running this function lots of times. What do you notice about the allele frequency over time?

```{r}
sim1 <- sim_freqs(N = 1000, mut_rate = 0)

```

Now repeat with a smaller population size (which represents the number of *infected* hosts in our case). What do you notice about the strength of drift? What happens once an allele reaches a frequency of 0 or 1? Why is this?

_**Definition:** Drift is the variation in the frequency of alleles in a small population due to the random loss of alleles across generations._

```{r}
sim_freqs(N = 100, mut_rate = 0)
```

Now repeat the simulation, this time running for 10 years and with some mutation. Do frequencies still always end up fixed/lost?

```{r}
sim_freqs(N = 1000, t_out = seq(0, 365 * 10), mut_rate = 1e-4)
```

### The allele frequency spectrum {-}

We will now simulate a large number of loci, but output at just a single timepoint.

```{r}
# simulate allele frequencies and look at the data
sim1 <- sim_freqs(N = 1000, t_out = 365 * 5, mut_rate = 1e-6, loci = 1e3, plot_on = FALSE)
head(sim1$data)
```

We can plot this allele frequency spectrum

```{r}
# simulate allele frequencies and look at the data
ggplot(sim1$data) +
  geom_histogram(aes(x = freq), breaks = seq(0, 1, 0.01))
```

What happens if you increase the mutation rate? What happens if you increase the population size?

### The buildup of homozygosity {-}

Run the code below multiple times to repeat the simulation of drift, but now also plotting homozygosity. What do you notice about how homozygosity changes over time? (It always increases to 1).

_**Definition:** Homozygosity is... ._

```{r}
sim_freqs(N = 1000, t_out = seq(0, 365 * 10, 7), plot_homo = TRUE)
```

What happens when you add in some mutation? Try mutation rates of 1e-3 and 1e-4. What happs to the equilibrium level of homozygosity?

```{r}
sim_freqs(N = 1000, t_out = seq(0, 365 * 10, 7), mut_rate = 1e-4, plot_homo = TRUE)
```

We can add in a line showing the theoretical expectation.

```{r}
sim_freqs(N = 1000, t_out = seq(0, 365 * 5, 7), mut_rate = 1e-4, plot_homo = TRUE, plot_expected_homo = TRUE)
```

## Multiple subpopulations {-}
Now we will consider an example with multiple subpopulations (demes).

### Perfectly isolated subpopulations {-}

Try running a simulation with 10 independent demes.This could represent ten villages where there is no movement of people between those villages. Notice that allele frequencies tend to diverge.

```{r}
sim1 <- sim_freqs(N = 1000, demes = 10, mig_rate = 0, mut_rate = 1e-4)
```

#### Calculating $F_{ST}$ to measure population differentiation{-}

**Definition: Fixation index, or $F_{ST}$ and Jost's D are both measures of genetic differentiation between different subpopulations.**

When $F_{ST}$ is close to **1** it means that individuals in a population tend to carry a _small number of alleles_. Individuals are “differentiated” in the sense that alleles are not freely mixing between people.



High values of $F_{ST}$ and Jost's D both suggest a greater genetic differentiation between the subpopulations being measured. However both metrics have different approaches to measure genetic differentiation. 


We can calculate a simple version of $F_{ST}$ manually

```{r}
sim_fst <- sim1$data %>%
  select(-locus) %>%
  group_by(time) %>%
  summarise(p_mean = mean(freq),
            p_var = var(freq),
            Fst = p_var / (p_mean * (1 - p_mean)))
```

Now lets plot $F_{ST}$ over time. What do you notice about differentiation?

```{r}
ggplot(sim_fst) +
  geom_line(aes(x = time, y = Fst)) +
  expand_limits(y = c(0, 1))
```

What happens if you repeat the process above with a higher mutation rate?

```{r}
sim1 <- sim_freqs(N = 100, demes = 10, mig_rate = 0, mut_rate = 1e-3,
                  t_out = seq(0, 365*10, 7), plot_on = FALSE)
sim1$data %>%
  select(-locus) %>%
  group_by(time) %>%
  summarise(p_mean = mean(freq),
            p_var = popvar(freq),
            Fst = p_var / (p_mean * (1 - p_mean))) %>%
  ggplot() + theme_bw() +
  geom_line(aes(x = time, y = Fst)) +
  expand_limits(y = c(0, 1))
```

The equilibrium level of $F_{ST}$is known to be 1/(1 + 2*N*mu) in this case. How does this level compare with what you saw in the plot?

```{r}
# e.g. use
#geom_hline(yintercept = 1 / (1 + 2*1e2*1e-3), linetype = "dashed")
```

The fact that $F_{ST}$ depends on the mutation rate can lead to problems. If one marker has a high mutation rate (e.g. microsatellites) then we would expect to see low values of $F_{ST}$, even if the populations have been evolving separately for a long time, and in this sense are well differentiated.

Jost's D is an alternative way of looking at differentiation.

When Jost's D is close to **1** it means that alleles tend to be present in _a small number of individuals_. Individuals are “differentiated” in the sense that they contain different genetic material.

```{r}
sim1 <- sim_freqs(N = 100, demes = 10, mig_rate = 0, mut_rate = 1e-3,
                  t_out = seq(0, 365*10, 7), plot_on = FALSE)

sim1$data %>%
  select(-locus) %>%
  group_by(time) %>%
  summarise(Js = mean(freq^2 + (1 - freq)^2),
            Jt = mean(freq)^2 + mean(1 - freq)^2,
            D = (Js - Jt) / Js * max(deme) / (max(deme) - 1)) %>%
  ggplot() + theme_bw() +
  geom_line(aes(x = time, y = D)) +
  expand_limits(y = c(0, 1.2))
```

Jost's D does not suffer from the same problem as $F_{ST}$, but it suffers from a different problem - it assumes infinite alleles mutation. When we have a biallelic SNP, Jost's D will tend to reach values of around 0.5 at equilibrium. (This is actually a bit of a subtle point - maybe easier to understand if I generalise the above to allow more than two alleles?)

In conclusion, both $F_{ST}$ and D have their limitations. $F_{ST}$ is good when mutation rates are low, for example biallelic SNPs. Jost's D is good for highly diverse and fast-mutating sites, such as microsats.


### Connected subpopulations {-}

What happens when we connect our demes up by migration? Try experimenting with larger and smaller migration rates in the following. What do you notice about the allele frequencies? (lines tend to be less bumpy and move together when migration is high)

```{r}
sim_freqs(N = 1000, demes = 5, mig_rate = 0.01)
```

Lets look at what happens to $F_{ST}$ when we include migration. What happens when you increase/decrease the migration rate?

```{r}
sim1 <- sim_freqs(N = 1000, demes = 10, t_out = seq(0, 365*20, 7), mig_rate = 1e-3,
                  mut_rate = 1e-6, plot_on = FALSE)
sim1$data %>%
  select(-locus) %>%
  group_by(time) %>%
  summarise(p_mean = mean(freq),
            p_var = popvar(freq),
            Fst = p_var / (p_mean * (1 - p_mean))) %>%
  ggplot() + theme_bw() +
  geom_line(aes(x = time, y = Fst)) +
  expand_limits(y = c(0, 1)) +
  geom_hline(yintercept = 1 / (1 + 2*1e3*(1e-3 + 1e-6)), linetype = "dashed")
```

Theory tells us that the equilibrium level of $F_{ST}$ in this situation is given by 1/(1 + 2*N*(m + mu)). Does this result hold up this your simulations? (yes, on average). Note that migration rates are usually much larger than mutation rates, meaning we can often simplify this to 1/(1 + 2*N*m).

Conclusion - at equilibrium, $F_{ST}$ reaches a level that is a balance between the strength of genetic drift (i.e. small population size) and migration. The higher the migration rate, the lower the value of $F_{ST}$. This means we can use $F_{ST}$ to tell us something about the connectivity of populations. Small pairwise $F_{ST}$ tends to indicate highly connected populations and vice versa.

Note, however, that these patterns take a very long time to build up. So we are not talking about the timescale of control programmes, but more of an evolutionary timescale.

-------------------------

# Analysis of DRC data {-}

Load the DRC MIP data. This data is a slightly simplified version of that used in this paper : [TO DO! LINK TO PAPER!!]

```{r, echo=FALSE, eval=FALSE}
# Code to process raw data from DRC paper. This code block does not need to be run by the user.

# read in raw data
dat <- readRDS(url("https://github.com/bobverity/antimalarial_resistance_DRC/raw/master/source_data/biallelic_distances.rds"))

# simplify
dat$distance <- NULL
dat$samples <- dat$samples %>%
  select(ID, Country, ADM1NAME, Year, lat, long)
names(dat$samples) <- c("ID", "Country", "Admin1_name", "Year", "Latitude", "Longitude")

dat$loci <- dat$loci %>%
  select(CHROM_NUMERIC, POS, REF, ALT, NEUTRAL, GEO) %>%
  mutate(NEUTRAL = c("Neutral", "Non-neutral")[match(NEUTRAL, c("True", "False"))],
         GEO = c("Geographic", "Non-geographic")[match(GEO, c("True", "False"))])
names(dat$loci) <- c("CHROM", "POS", "REF", "ALT", "NEUTRAL", "GEO")

# get unique lat-lon into clusters
latlon <- interaction(round(dat$samples$Latitude, 2), round(dat$samples$Longitude, 2))
dat$samples$Cluster <- as.factor(match(latlon, unique(latlon)))

# save to file
saveRDS(dat, file = "data/DRC_MIPs_biallelic_processed.rds")
```

Have a look at the data using the *MIPanalyzer* package.

```{r}
# load data
MIP_data <- readRDS("data/DRC_MIPs_biallelic_processed.rds")

# have a look at the data
MIP_data

head(MIP_data$samples)

head(MIP_data$loci)
```

let's create a copy of the dataset that does not include any DRC samples:

```{r}
# filter samples
MIP_data_noDRC <- MIPanalyzer::filter_samples(MIP_data, MIP_data$samples$Country != "DRC")

MIP_data_noDRC
```

PCA - [to do - explain what a PCA is ]


Now lets do PCA on the within-sample allele frequencies

```{r}
wsaf_impute <- MIPanalyzer::get_wsaf(MIP_data_noDRC, impute = TRUE, FUN = mean)
pca <- MIPanalyzer::pca_wsaf(wsaf_impute)

plot_df <- data.frame(PC1 = pca$x[,1],
                      PC2 = pca$x[,2],
                      Country = MIP_data_noDRC$samples$Country)

ggplot(plot_df) + theme_bw() +
  geom_point(aes(x = PC1, y = PC2, color = Country))
```

We see well-separated clusters, which makes sense geographically. Now let's repeat the process using all samples including those from DRC.

```{r}
wsaf_impute <- MIPanalyzer::get_wsaf(MIP_data, impute = TRUE, FUN = mean)
pca <- MIPanalyzer::pca_wsaf(wsaf_impute)

plot_df <- data.frame(PC1 = pca$x[,1],
                      PC2 = pca$x[,2],
                      Country = MIP_data$samples$Country)

ggplot(plot_df) + theme_bw() +
  geom_point(aes(x = PC1, y = PC2, color = Country))
```

The DRC points connect the other points. This suggests that there is more of a spatial continuum of allele frequencies.

We can plot the contributions that each locus makes to this PCA. Starting with PC1, what do you notice about the balance of geographically informative vs. non-informative SNPs? (the pattern is mostly driven by geographically informative SNPs). What does this suggest? (the primary factor driving this PCA is spatial location)

```{r}
# plot component 1
MIPanalyzer::plot_pca_contribution(pca, component = 1, chrom = MIP_data$loci$CHROM, pos = MIP_data$loci$POS,
                                   locus_type = MIP_data$loci$GEO)
```

Now repeat the process with PC2. What do you notice about where these loci are on the genome? (they are all in one or two places). What genes are at these positions? (pfcrt on chrom7. Small peak at dhps on chromosome 8)

```{r}
# plot component 2
MIPanalyzer::plot_pca_contribution(pca, component = 2, chrom = MIP_data$loci$CHROM, pos = MIP_data$loci$POS,
                                   locus_type = MIP_data$loci$GEO)
```


-------------------------

# Inferring population structure using *rmaverick* {-}


[explain what *rmaverick* is and how going to use it ]

Load barcode data from Amy Bei's 2018 paper [link to paper]. This data has already been filtered to remove the two Thies samples, and to order in terms of date (2001-2002 vs. 2014) and location (Dielmo and Ndiop)

```{r, echo=FALSE, eval=FALSE}
# code not run - this is how I get the raw data from SIMPLEGEN and process it
load("/Users/rverity/Dropbox/Bob/Work/My Programs/Simulation/SIMPLEGEN/data/Bei_2018.RData")
Bei_2018_processed <- Bei_2018

Bei_2018_processed$barcodes <- Bei_2018_processed$barcodes %>%
  filter(Location %in% c("Dielmo", "Ndiop")) %>%
  mutate(Date = as.Date(Date, format = "%m/%d/%Y"),
         Date_level = (Date > as.Date("2010-01-01")) + 1) %>%
  arrange(Date_level, Location, Date) %>%
  select(-Haplotype, -Date_level, -`Sample ID`, -COI) %>%
  as.data.frame()

Bei_2018_processed$EIR <- as.data.frame(Bei_2018_processed$EIR)

Bei_2018_processed$SNP_locations <- as.data.frame(Bei_2018_processed$SNP_locations)

save(Bei_2018_processed, file = "data/Bei_2018_processed.RData")
```

 Now let's look at a snapshot of the data, which includes measures of transmission intensity in the two timepoints and regions. We will look at:
 - Entomological Innoculation Rate (EIR)
 - The barcodes
 - The SNP locations

```{r}
load("data/Bei_2018_processed.RData")

head(Bei_2018_processed$EIR)

head(Bei_2018_processed$barcodes)

head(Bei_2018_processed$SNP_locations)
```

Some data processing

```{r}
# filter out polygenomics
mav_data <- Bei_2018_processed$barcodes %>%
  filter(`M/P genomic` == "M")

# convert genotype calls to numeric
for (i in 8:31) {
  mav_data[,i] <- match(mav_data[,i], c("A", "C", "T", "G"))
}

head(mav_data)
```

Next, load the data into *rmaverick*, and make a first model set [explain]. Print the project to check everything looks as planned (25 samples, 24 loci etc.)

```{r}
myproj <- rmaverick::mavproject() %>%
  rmaverick::bind_data(df = mav_data, ID_col = 1, data_cols = 8:31, ploidy = 1) %>%
  rmaverick::new_set(name = "no admixture model", admix_on = FALSE)

myproj
```

[explain what MCMC is]

Run MCMC. Users should run this with pb_markdown = FALSE.

```{r}
myproj <- rmaverick::run_mcmc(myproj, K = 1:8, burnin = 1e3, samples = 1e3,
                              rungs = 10, GTI_pow = 1.5, pb_markdown =  TRUE)

```

Some MCMC diagnostic checks. First, check that coupling rates are above 0. Do this for every value of K from 2 to the max value explored.

[define coupling rates and definition]

```{r}
rmaverick:::plot_mc_acceptance(myproj, K = 8)
```

[define what k parameter does]

Now check what the program thinks about the best value of K. In this case, values around 3-5 are reasonably likely

```{r}
rmaverick::plot_logevidence_K(myproj)
rmaverick::plot_posterior_K(myproj)
```

Produce Structure plot. [explain what this is] Add vertical lines to break up into the following groups:
1. Dielmo 2001-2001
2. Ndiop  2001-2002
1. Dielmo 2014
2. Ndiop  2014

```{r}
rmaverick::plot_qmatrix(myproj, K = 3:5) +
  geom_vline(xintercept = c(4, 10, 16) + 0.5, size = 1)
```



Notice that the later samples are (mostly) clearly allocated to one or other population with high probability. These essentially represent clonal lineages. We can see that groups 2,3,4 are present in Dielmo in 2014, and in Ndiop we additionally have group 1.

Comparing this to the earlier time point, samples are far more ambigious in their allocation. They look a bit like some of these clonal lineages, but it's not clear cut, probably because there is some additional variation. Interestingly, group 4 dominates in Ndiop in 2001-2002. This lineage is almost completely lost by the later sampling point.

Overall, we can see evidence of increasing population structure over time. Compare this with the EIR values - what do you notice? (lower transmission later in time, therefore increasing differentiation and population structure).

-------------------------

Concluding questions to summarize:
1.
2.
3.