---
title: "AMMS Population Structure from Allele Frequencies Practical" 
author: "Izzy Routledge, Bob Verity"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    theme: readable
    highlight: tango
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, 
                      fig.align = 'center', fig.keep = 'all')
```

<!-- Here we style out button a little bit -->

```{=html}
<style>
.showopt {
background-color: #004c93;
color: #FFFFFF; 
width: 100px;
height: 20px;
text-align: center;
vertical-align: middle !important;
border-radius: 8px;
float:right;
}

.showopt:hover {
background-color: #dfe4f2;
color: #004c93;
}

</style>
```

# Dependencies for Practical {-}

Please copy and paste the below code chunk in it's entirety to your console to download R package libraries needed for this practical. If you are having trouble installing any of the R packages, please ask an instructor for a pre-loaded flash drive.

```{r, echo=T}
# deps <- c("tidyverse", "vcfR", , "sf", "tidygraph", "hierfstat","rmaverick")
# deps <- !sapply(deps, function(x){x %in% installed.packages()[,1]} )
# if(any(deps)) {0
#   if(deps["hierfstat"]) {
#     if(!"gaston" %in% installed.packages()[,1]){
#       install.packages("gaston")
#     }
#     }if(deps["rmaverick"]){
#       devtools::install_github("bobverity/rmaverick")
#     } else  {
#     install.packages(names(deps)[deps])
#   }
# }


```

Now load all of those libraries into this session using the code chunk below. Please copy and paste it in its entirety.

```{r, echo=T}
library(tidyverse)
library(vcfR)
library(rmaverick)
library(hierfstat) 
library(sf)
library(tidygraph)
```

# Intro to population structure and allele frequency metrics {-}
-Why useful for epi?
***

# Overview of Data {-}
 
***

# Practical Goals {-}
By the end of this practical, you should be able to: 

- Calculate allele frequencies from genetic data 
- Estimate metrics such as FST and explain what they can tell us about population differentiation
- Appreciate how allele frequency methods can identify connectivity over large spatial scales and long temporal scales
- Explain how allele frequencies can give useful information, and also describe their limitations for malaria control purposes


-------------------------
# Importing Data

Here, we will use the `vcfR` [package](https://knausb.github.io/vcfR_documentation/) to read in our variant call file (VCF) that contains biallelic SNPs for our \*\*\* individuals across our four populations.

Read in the VCF from our data folder that you downloaded from Github: `data/INSERTFILEPATH.vcf`. If you are having trouble locating the file on your computer, please type `file.choose` into your console for an interactive option to locate the file. Name the object (the vcf file you are reading in) *vcf* for simplicity.

```{r}
#vcf <- vcfR::read.vcfR("data/FILEPATH.vcf")
```



# Calculating Allele Frequencies from VCF data
- Define allele frequencies
- Explain 
- Recycle code from drug resistance practical 

```{r}


```
# Calculating FST
- Define FST
- Explain how and when it is used
- recycle code from Namibia? But not in VCF format

```{r}


```

# Using PCA / PCoA to identify population structure




```{r}


```
#STRUCTURE / rmaverick



```{r}


```
#Tests for  isolation by distance (e.g. Rousset’s method, Mantel test)


```{r}


```