---
title: "Dependencies"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Dependencies for Workshop
Software dependenices are defined as pieces of code or software that depend on one another. In our case, we will be using specific R-packages to help power the practicals. Some of these R-packages are availabe from [CRAN](https://cran.r-project.org/), a common R package repository, or [Github](https://github.com/), a common place for open-source code. Below you will find instructions on how to install these various dependencies. 

## CRAN Packages
Please copy and paste each line of code below into your console. *It is important to copy one line of code at a time.* If you run into problems downloading a specific package, please make a note of it and notify an instructor at an appropriate time. 
```{r, eval = F}
install.packages("tidyverse")
install.packages("vcfR")
install.packages("sf")
install.packages("tidygraph")
install.packages("ggraph")
install.packages("cowplot")
install.packages("remotes") # using this package below 
```

## Github Packages
As above, please copy and paste each line of code below into your console. *Again, it is important to copy one line of code at a time.* If you run into problems downloading a specific package, please make a note of it and notify an instructor at an appropriate time. 
```{r, eval = F}
remotes::install_github("bobverity/rmaverick")
remotes::install_github("OJWatson/hmmibdr")
remotes::install_github("mrc-ide/MIPanalyzer")
```
