---
title: "AMMS COI Practical" 
author: "Jason Hendry, Nick Brazeau"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  html_document:
    theme: readable
    highlight: tango
    code_folding: hide
    toc: yes
    toc_float:
      collapsed: false
      smooth_scroll: false
    toc_depth: 2
    number_sections: true
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, warning = F, message = F, 
                      fig.align = 'center', fig.keep = 'all')
```

<!-- Here we style out button a little bit -->

```{=html}
<style>
.showopt {
background-color: #004c93;
color: #FFFFFF; 
width: 100px;
height: 20px;
text-align: center;
vertical-align: middle !important;
border-radius: 8px;
float:right;
}

.showopt:hover {
background-color: #dfe4f2;
color: #004c93;
}

</style>
```
<!--Include script for hiding output chunks-->

```{=html}
<script src="misc/hideOutput.js"></script>
```
# Dependencies for Practical {.unnumbered}

Please copy and paste the below code chunk in it's entirety to your console to download R package libraries needed for this practical. If you are having trouble installing any of the R packages, please ask an instructor for a pre-loaded flash drive.

```{r, echo=T}
deps <- c("tidyverse", "vcfR")
deps <- !sapply(deps, function(x){x %in% installed.packages()[,1]} )
if(any(deps)) {
    install.packages(names(deps)[deps])
}

```

Please now load all of those libraries into this session using the code chunk below. Please copy and paste it in its entirety.

```{r, echo=T}
library(tidyverse)
library(vcfR)
```

# Practical Goals {.unnumbered}

By the end of this practical, you should understand the following concepts:  
- ***

------------------------------------------------------------------------

# Importing Data

Here, we will use the `vcfR` [package](https://knausb.github.io/vcfR_documentation/) to read in our variant call file (VCF) that contains biallelic SNPs for our *** individuals across our four populations.

```{r}
drc <- vcfR::read.vcfR("data/SNPs.DRCongo.setB.random96.vcf")
vnm <- vcfR::read.vcfR("data/SNPs.Vietnam.setD.random96.vcf")
```
